---
title: "FunctionAnalysis"
author: "XWen"
date: "June 29, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy = TRUE)
```

## GENERATE DEG VECTOR

```{r importData & select up-gragulated genes & sorted by logFC}
# import dataset
DGE_padjSorted <- read.delim("DGE_padjSorted.txt")
dim(DGE_padjSorted)

# rownames to column as Gene
library(tibble)
DGE_padjSorted <- tibble::rownames_to_column(DGE_padjSorted, "Gene")

# select logFC > 1.5 & extract logFC column
DGE_FilteredSorted <- DGE_padjSorted[DGE_padjSorted$logFC > 1.5, 1:2]
# sorted according to  logFC 
DGE_FilteredSorted <- DGE_FilteredSorted[order(DGE_FilteredSorted$logFC, decreasing = TRUE),]

# set gene as rowname
rownames(DGE_FilteredSorted) <- DGE_FilteredSorted[,1]
# drop gene column
DGE_FilteredSorted <- subset(DGE_FilteredSorted, select=-Gene)

# check dimension
dim(DGE_FilteredSorted)
```


##### add ENTREZ IDs to the DGE result
```{r loadAnnotationLibrary}
library(AnnotationDbi)
library(org.Hs.eg.db)
keytypes(org.Hs.eg.db)
```

```{r addAnnotation ENTREZID}
# select()returned 1:1 mapping between keys and columns
annots <- select(org.Hs.eg.db,keys=rownames(DGE_FilteredSorted),
                 columns=c("SYMBOL","ENTREZID"),keytype="SYMBOL") 
# merge annotation with logFC df
resultTable <- merge(DGE_FilteredSorted,annots,by.x=0,by.y="SYMBOL")
# set gene as row name
rownames(resultTable) <- resultTable[,1]
# drop gene column
resultTable <- subset(resultTable,select=-Row.names)
resultTable <- resultTable[order(resultTable$logFC, decreasing = TRUE),]
head(resultTable)
```

```{r save results}
# save the results
write.table(annots, file = "annotation_logFC_filtered.txt",quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
write.table(resultTable, file = "dgewAnno_logFC_filtered.txt",quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
write.table(DGE_FilteredSorted, file = "DGElogFC_filtered.txt",quote = FALSE, sep = "\t", row.names = TRUE, col.names = TRUE)
```


## FUNCTIONAL ENRICHMENT ANALYSIS
```{r fx enrichment analysis}
library(clusterProfiler)
# Gene Ontology that the up-regulated DEG are involved in
# cellular components 
enrich_cc <- enrichGO(annots$ENTREZID, keyType = "ENTREZID",
                      OrgDb = org.Hs.eg.db,
                      ont = "CC", 
                      readable = TRUE)

# molecular functions 
enrich_mf <- enrichGO(annots$ENTREZID, keyType = "ENTREZID",
                      OrgDb = org.Hs.eg.db,
                      ont = "MF", 
                      readable = TRUE)

# biological processes 
enrich_bp <- enrichGO(annots$ENTREZID, keyType = "ENTREZID",
                      OrgDb = org.Hs.eg.db,
                      ont = "BP", 
                      readable = TRUE)
```

##### Visualization of Results

```{r enrichGO barplot CC}
# up-regulated DGE genes involved significant cellular components
barplot(enrich_cc)

pdf("up_gene_CellComponent.pdf")
barplot(enrich_cc)
dev.off()
```

```{r enrichGO barplot BP}
# up-regulated DGE genes involved significant biological process
barplot(enrich_bp)

pdf("up_gene_BioProcess.pdf")
barplot(enrich_bp)
dev.off()
```

```{r enrichGO barplot MF}
# up-regulated DGE genes involved significant molecular function
barplot(enrich_mf)

pdf("up_gene_molecularfx.pdf")
barplot(enrich_mf)
dev.off()
```


##### KEGG (Kyoto Encyclopedia of Genes and Genomes) Analysis

```{r enrichKEGG dotplot}
enrich_kegg <- enrichKEGG(annots$ENTREZID)
dotplot(enrich_kegg)

pdf("up_gene_keggPathways.pdf")
dotplot(enrich_kegg)
dev.off()
```

## GENE-CONCEPT NETWORK
```{r enrichedKEGG involved gene network}
# to know which genes are involved in those terms
enrich_keggGene <- setReadable(enrich_kegg,OrgDb = org.Hs.eg.db,keyType = "ENTREZID")

# visualize network
cnetplot(enrich_keggGene,foldChange = DGE_FilteredSorted$logFC, categorySize='pvalue',colorEdge = TRUE)

pdf("geneNetwork_enrichKEGG.pdf")
cnetplot(enrich_keggGene,foldChange = DGE_FilteredSorted$logFC, categorySize='pvalue',colorEdge = TRUE)
dev.off()
```

## GLOBAL/ UNIVERSAL GENE SET ENRICHMENT ANALYSIS (GSEA)

```{r GSEA - download GSEA-MsigDB H set}
# download <H: hallmark gene sets collection> from GSEA-MsigDB
library(msigdbr)

# obtain gene set
msig <- msigdbr(species="Homo sapiens", category="H")

# select desired information
h <- msig %>% select (gs_name, entrez_gene)
```

##### Generating Global DEG Vector including all of the DGE sorted by logFC (in descending order) named with ENTREZIDs.
```{r load dge result & sorted by logFC, & extract logFC}
DGE_sorted <- read.delim("DGE_padjSorted.txt")

# order by logFC
DGE_sorted <- DGE_sorted[order(DGE_sorted$logFC, decreasing = TRUE),]

# subset of DGE_sorted by selecting logFC
DGE_sorted <- subset(DGE_sorted,select=logFC)
dim(DGE_sorted)
```

##### add entrezid to the df
```{r load annotationDbi library}
library(AnnotationDbi)
library(org.Hs.eg.db)
keytypes(org.Hs.eg.db)
```

```{r convert symbol to enterezid, merge with logFC df & extract sorted logFC, named with entrezid}
# select()returned 1:1 mapping between keys and columns
annots_all <- AnnotationDbi::select(org.Hs.eg.db, keys=rownames(DGE_sorted),
                                    columns=c("SYMBOL","ENTREZID"),
                                    keytype="SYMBOL") 

# merge annotation with logFC df
resultTable_all <- merge(DGE_sorted,annots_all,by.x=0,by.y="SYMBOL")

# rename column names
names(resultTable_all) <- c("Gene", "logFC", "ENTREZID")

# drop na and duplicates of entrezid
library(tidyr)
library(dplyr)
resultTable_all <- resultTable_all%>% drop_na(ENTREZID) %>% distinct(ENTREZID, .keep_all = TRUE)


# # set entrezid as row name
# rownames(resultTable_all) <- resultTable_all[,3]
# # drop entrezid column
# resultTable_all <- subset(resultTable_all,select=-ENTREZID)
# # sorted by logFC
# resultTable_all <- resultTable_all[order(resultTable_all$logFC, decreasing = TRUE),]
# head(resultTable_all)


# # drop na and duplicates of gene
# library(tidyr)
# library(dplyr)
# resultTable_all <- resultTable_all%>% drop_na(Gene) %>% distinct(Gene, .keep_all = TRUE)



# sorted by logFC 
resultTable_all <- resultTable_all[order(resultTable_all$logFC, decreasing = TRUE),]

# extract the logFC
result_logFC <- resultTable_all$logFC
# name the vector
names(result_logFC) <- resultTable_all$ENTREZID
 
result_logFC[1:5]
```

#### GSEA analysis
```{r gsea analysis of the sorted logFC named by entrezid}
# Perform GSEA Analysis
gsea_res <- GSEA(result_logFC, TERM2GENE=h)
```


```{r gsea geneset id}
# view gene set ID
gsea_res@result$ID
```

```{r gsea plot 1:5}
# view gsea plot
library(enrichplot)
gseaplot2(gsea_res,geneSetID = 1:5 )

pdf("gse_plot.pdf")
gseaplot2(gsea_res,geneSetID = 1:5 )
dev.off()
```

```{r gsea plot 22:27}
gseaplot2(gsea_res,geneSetID = 22:27 )
pdf("gse_plot2.pdf")
gseaplot2(gsea_res,geneSetID = 22:27 )
dev.off()
```

#### TRANSCRIPTION FACTOR ANALYSIS

```{r download GSEA-MsigDB C3 set: regulatory gene set}
# download <C3: regulatory gene sets collection> from GSEA-MsigDB
library(msigdbr)

# obtain gene set
msigc3 <- msigdbr(species="Homo sapiens", category="C3")

# select desired information
c3 <- msigc3 %>% select (gs_name, entrez_gene)
```


```{r enrich up-regulated c3 gene set collection}
upEntrezid_c3_enricher <- enricher(resultTable$ENTREZID,TERM2GENE = c3)

# Convert ENTREZIDs to Gene Symbols
upGene_c3_enricher <- setReadable(upEntrezid_c3_enricher,OrgDb=org.Hs.eg.db, keyType = "ENTREZID")

```


```{r visual up-enriched regulatory gene network}
cnetplot(upGene_c3_enricher,foldChange=resultTable$logFC, categorySize="pvalue", colorEdge = TRUE)

pdf("geneNet_transcriptionFactors_4upGenes.pdf")
cnetplot(upGene_c3_enricher,foldChange=resultTable$logFC, categorySize="pvalue", colorEdge = TRUE)
dev.off()
```

