---
title: "MacroArrayQC"
author: "XWen"
date: "June 23, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy = TRUE)
```

## QC analysis for macroarray dataset

Data source: GSE32323 and GSE8671
##### import dataset 
```{r cars}
library(affy)
raw32323 <- ReadAffy(celfile.path="GSE32323_RAW")
raw8671 <- ReadAffy(celfile.path = "GSE8671_RAW")
```


```{r}
# print sample names for GSE32323
colnames(raw32323@assayData$exprs)
```


```{r}
# print sample names for GSE8671
colnames(raw8671@assayData$exprs)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
# briefly view the head rows 
# df32323 <- as.data.frame(exprs(raw32323))
# View(head(df32323))
```

```{r}
# df8671 <- as.data.frame(exprs(raw8671))
# View(head(df8671))#
```

### QC analysis 
##### arrayQualityMetrics package
```{r}
library(arrayQualityMetrics)
```

```{r}
arrayQualityMetrics(raw32323,
  outdir = 'raw32323_QCReport',
  force = TRUE,
  do.logtransform = TRUE)
```

```{r}
arrayQualityMetrics(raw8671,
  outdir = 'raw8671_QCReport',
  force = TRUE,
  do.logtransform = TRUE)
```


##### affyPLM package
```{r}
library(affyPLM)
```

```{r}
Pset32323 <- fitPLM(raw32323)

pdf(file = "Pset32323_RLE.pdf", width = 6, height = 4)
RLE(Pset32323)
dev.off()

pdf(file = "Pset32323_NUSE.pdf", width = 6, height = 4)
NUSE(Pset32323)
dev.off()
```


```{r}
Pset8671 <- fitPLM(raw8671)

pdf(file = "Pset8671_RLE.pdf", width = 6, height = 4)
RLE(Pset8671)
dev.off()

pdf(file = "Pset8671_NUSE.pdf", width = 6, height = 4)
NUSE(Pset8671)
dev.off()
```

## background correction and normalization ####

```{r}
exprset32323 <- rma(raw32323)
write.exprs(exprset32323, file="exprset32323.txt")

exprset8671 <- rma(raw8671)
write.exprs(exprset8671, file="exprset8671.txt")
```

```{r}
par(mfrow=c(1,2))
boxplot(raw32323)
hist(raw32323)

# pdf(file = "boxplot_gse32323.pdf", width = 6, height = 4)
# boxplot(raw32323)
# dev.off()
```

```{r}
par(mfrow=c(1,2))
boxplot(raw8671)
hist(raw8671)

# pdf(file = "boxplot_gse8671.pdf", width = 6, height = 4)
# boxplot(raw8671)
# dev.off()
```


```{r}
phenoData(raw32323)
pData(raw32323)
```

```{r}
phenoData(raw8671)
pData(raw8671)
```


## BATCH CORRECTION XXXXXXXXXXXXXXXXXXXXXXXXXXX
```{r}
library(sva)
```


```{r}
# pheno input
pheno <- rbind.data.frame(pData(exprset32323), pData(exprset8671))
pheno['group'] <- c(rep(c("Normal","Cancer"),each=17),
                    rep(c("5aza_CancerCell","cancerCell"),times=5),
                    rep(c("Normal","Cancer"), each=32))
pheno['batch'] <- c(rep(c("A","B"),c(44,64)))
```


```{r}
# edata input 
edata = cbind(exprs(exprset32323),exprs(exprset8671))
```

```{r}
# create the full model matrix
mod = model.matrix(~as.factor(group), data=pheno)
#batch variable
batch <- pheno$batch
```

```{r}
combat_edata = ComBat(dat=edata, batch=batch, mod=mod)

```

```{r}
dim(combat_edata)
write.table(combat_edata, file = "combat_edata.txt")
```

```{r}
# convert affymatrix proble into gene symbols
library(biomaRt)

probes <- row.names(combat_edata)

mart <- biomaRt::useEnsembl(biomart="ensembl",
                            dataset="hsapiens_gene_ensembl")

symbol_id_prob <- biomaRt::getBM(attributes=c("hgnc_symbol", "ensembl_gene_id",
                            "affy_hg_u133_plus_2"),
               filters = "affy_hg_u133_plus_2",
               values = probes,
               mart = mart)

head(symbol_id_prob)
```

```{r}
dim(symbol_id_prob)
```


```{r}
row.names(combat_edata)[1:5]
```

```{r}
# drop duplicates
symbol_id_prob <- as.data.frame(symbol_id_prob %>% group_by(affy_hg_u133_plus_2) %>% slice(1))

# set row name
rownames(symbol_id_prob) <- symbol_id_prob$affy_hg_u133_plus_2

dim(symbol_id_prob)
```


```{r}
#Combine gene annotations with raw data
combat_edata_symId=merge(symbol_id_prob,combat_edata, by="row.names", all = TRUE)

#Write GCRMA-normalized, mapped data to file
write.table(combat_edata_symId, file = "combat_edata_symId.txt", quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
```


### read data ######################################################
```{r}
# combat_edata_symId <- read.delim("C:/Users/wen_x/Downloads/R/Bioinfo/M3/combat_edata_symId.txt")
```



```{r}
library(dplyr)
# calculate median value for each gene
combat_edata_symMedian <- combat_edata_symId  %>%  group_by(hgnc_symbol) %>% 
  summarise_at(vars(colnames(combat_edata_symId[,5:112])),median)
# drop rows with hgnc_symbo = na
library(tidyr)
combat_edata_symMedian <- as.data.frame(combat_edata_symMedian %>% drop_na(hgnc_symbol))

# use hgnc_symbl as row names
rownames(combat_edata_symMedian) <- combat_edata_symMedian$hgnc_symbol
```

```{r}
combat_edata_symMedian <- combat_edata_symMedian[,2:109]
combat_edata_symMedian[1:3,1:3]
```


```{r}
## DATA VISUALIZATIONS
##### boxplot

boxplot(combat_edata_symMedian)

pdf(file = "boxplot_combat_edata_symMedian.pdf")
boxplot(combat_edata_symMedian)
dev.off()
```

##### PCA
```{r}
pca <- prcomp(t(combat_edata_symMedian), scale = FALSE, center = FALSE)

```

```{r}
head(pca$x[,1:5])
```


```{r}
library(ggplot2)
```
```{r}
library(dplyr)
pheno2 <- data.frame(pheno, batch_group = paste("batch",pheno$batch, pheno$group, sep = "_"))
df <- merge(pheno2,pca$x[,1:2], by.x = 0, by.y = 0)
```

```{r}
dim(df)
```
```{r}
plot(pca$x[,1], pca$x[,2])
```
```{r}
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent
Variation")
```


```{r}
ggplot(df, aes(PC1, PC2, color=batch_group, label=sample))+
  geom_point()+
  geom_text(size = 2.5, position = position_dodge(width = 1),
    vjust = -0.5)+
  stat_ellipse()+
  xlab(paste("PC1 - ", pca.var.per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca.var.per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PCA Graph")
  

```

```{r}
pdf(file = "PCA_combat_edata.pdf")
ggplot(df, aes(PC1, PC2, color=batch_group, label=sample))+
  geom_point()+
  geom_text(size = 2.5, position = position_dodge(width = 1),
    vjust = -0.5)+
  stat_ellipse()+
  xlab(paste("PC1 - ", pca.var.per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca.var.per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PCA Graph")
  
dev.off()

```

```{r}
## get the name of the top 10 measurements (genes) that contribute
## most to pc1.
loading_scores <- pca$rotation[,1]
gene_scores <- abs(loading_scores) ## get the magnitudes
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)
top_10_genes <- names(gene_score_ranked[1:10])
 
top_10_genes ## show the names of the top 10 genes
 

```
```{r}
pca$rotation[top_10_genes,1] ## show the scores (and +/- sign)
```


#####Correlation Heatmaps
```{r}
library(pheatmap)
```


```{r}
# hierarchical clustering in the same manner as performed by pheatmap to obtain gene clusters
library(dendextend)
hclust_gene <- hclust(dist(combat_edata_symMedian), method = "complete")

as.dendrogram(hclust_gene) %>%
  plot(horiz = TRUE)
```



```{r}
# cut tree for k = 3
gene_row <- cutree(tree = as.dendrogram(hclust_gene), k = 3)
gene_row <- as.data.frame(gene_row) 
head(gene_row)
```

```{r}
# add random numbers 1:3
set.seed(1000)
thr_random <- as.factor(sample(x = 1:3, size = nrow(gene_row), replace = TRUE))
gene_row$random <- thr_random
head(gene_row)
```

```{r}
sample_col <- cbind.data.frame(df$Row.names,df$group)
rownames(sample_col) <- sample_col$`df$Row.names`
colnames(sample_col) <- c('rownames','group')
sample_col <- subset (sample_col, select = group)
head(sample_col)
```


```{r}
pheatmap(combat_edata_symMedian, annotation_row = gene_row, annotation_col = sample_col, 
         show_rownames = FALSE, show_colnames = FALSE)
```

```{r}
pdf(file="pheatmap2.pdf")
pheatmap(combat_edata_symMedian, annotation_row = gene_row, annotation_col = sample_col, 
         show_rownames = FALSE, show_colnames = FALSE)
dev.off()
```


